\documentclass[]{scrreprt}
\usepackage{amsmath,amsfonts,graphicx}
\usepackage{multirow}
\usepackage{pslatex}
\usepackage{tabularx}
\usepackage{comment}
\usepackage{xspace}
\usepackage{array}

\usepackage{hyperref}

\usepackage{caption}
\DeclareCaptionFont{white}{\color{white}}
\DeclareCaptionFormat{listing}{\colorbox{gray}{\parbox{\textwidth}{#1#2#3}}}

\graphicspath{
{figures/}
}

\def\species{\mathrm{sp}}
\def\phase{\mathrm{ph}}
\def\massfrac{\chi}
\def\flux{\mathbf{F}}
\def\darcyvel{\mathbf{v}}
\def\energydens{\mathcal{E}}
\def\d{\mathrm{d}}

\newcommand{\uo}{\mbox{UO\textsubscript{2}}\xspace}

\setcounter{secnumdepth}{3}


\begin{document}


\title{Jacobian Tests}
\author{CSIRO}
\maketitle

\tableofcontents

\chapter{Jacobian Tests}

\section{Fluid-mass time derivative}

The following tests of the Jacobian are performed on the fluid-mass time derivative kernel (PorousFlowMassTimeDerivative).
\begin{enumerate}
\item single-phase, with a single component, with a van-Genuchten capillary pressure, constant bulk-modulus density and constant porosity.
\item single-phase, with 3 components, with a van-Genuchten capillary pressure, constant bulk-modulus density and constant porosity.
\item 2-phase PP formulation, with 2 components (that exist in both phases), with a van-Genuchten capillary pressure, constant bulk-modulus density for each phase and constant porosity.
\item 2-phase PP formulation, with 3 components (that exist in both phases), with a van-Genuchten capillary pressure, constant bulk-modulus density for each phase and constant porosity.
\end{enumerate}


\section{Fluid Advective Flux}

The following tests of the Jacobian are performed on the fluid-mass advective flux kernel (PorousFlowAdvectiveFlux).
\begin{enumerate}
\item single-phase, with 1 component, constant viscosity, constant insitu-permeability, density with constant bulk modulus, Corey relative permeability, nonzero gravity, with van-Genuchten capillary pressure.
\item single-phase, with 3 components, constant viscosity, constant insitu-permeability, density with constant bulk modulus, Corey relative permeability, nonzero gravity, with van-Genuchten capillary pressure.
\item 2-phase (PP formulation), with 2 components (that exist in both phases), constant viscosity for each phase, constant insitu-permeability, density with constant bulk modulus for each phase, Corey relative permeability for each phase, nonzero gravity, with van-Genuchten capillary pressure.
\item 2-phase (PP formulation), with 3 components (that exist in both phases), constant viscosity for each phase, constant insitu-permeability, density with constant bulk modulus for each phase, Corey relative permeability for each phase, nonzero gravity, with van-Genuchten capillary pressure.
\end{enumerate}

\end{document}

